---
title: "00_demo"
author: "Danica Schaffer-Smith"
date: "May 12, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load packages
```{r pkgs}
library(SWATplusR)
```


## Set up SWAT Project Directory and Demo Projects

The R package SWATdata provides a set of fast running, lightweight SWAT2012 and 
SWAT+ model setups of a head watershed of the Little River Experimental 
Watershed (LREW; Bosch et al., 2007). Additionally, discharge observations at 
the outlet of the demo catchment (Gauge J of LREW) and spatial information of 
the SWAT model setups are available from SWATdata. An overview of the available 
data sets is available on the SWATdata GitHub page.

The demo data that is available from SWATdata can be loaded with the function 
load_demo(). Loading an entire SWAT project to your hard drive requires the 
information that the loaded dataset is a 'project'. As the project will be 
written to the hard drive a path is required. Further, the SWAT version must be 
provided, to either load a SWAT+ ('plus'), or a SWAT2012 ('2012') project. 
Optionally, the revision number can be provided if a specific model revision 
(only for SWAT+) should be loaded. An overview of the different SWAT demo 
project setups (with their revision numbers) is available at the SWATdata GitHub 
page. If no revision number is provided, a project with the most recent revision 
number available fom SWATdata is loaded. A demo project can be loaded with the 
following command:

```{r swat dir}
# choose the path where the SWAt demo project will be stored
demo_path <- 'data_input/SWAT_demo'

# Load SWAT2012 demo project
# load_demo() loads a demo project dataset and saves it in the demo_path.
# This path can be used to run the swat model located at the path.
path_2012 <- load_demo(dataset = "project",
                       version = "2012",
                       path = demo_path)

# Loading a SWAT+ demo project
path_plus <- load_demo(dataset = "project",
                       version = "plus",
                       path = demo_path,
                       revision = 57)

```

## Observation data

SWATplusR provides observation data of daily discharge records at the main outlet 
of the demo catchment (Gauge J of LREW) for the time period 1968-01-01 through
2012-12-31. The data set is used for model evaluation. 

Load observed data as follows:

```{r pressure, echo=FALSE}
q_obs <- load_demo(dataset = "observation")

q_obs
```

Create a quick plot of the observed flow data
```{r plot q, echo=FALSE}
plot(q_obs, type = "l")
```

## Spatial data

To visualize simulation results spatially, shape files of the subbasins, river 
network, and HRUs for the SWAT+ and the SWAT2012 demo setups are available from 
SWATplusR. For spatial data load_demo() only returns the paths to the shape 
files. As the spatial data for the two SWAT project setups looks different 
loading the data requires the definition of the SWAT version.

```{r spatial}
# Path to the subbasin shape file
sub_path <- load_demo(dataset = "subbasin", version = "plus")
# Path to the stream network shape file
riv_path <- load_demo(dataset = "river", version = "plus")

sub_path
#[1] "C:/Users/Schaffer-Smith/Documents/R/win-library/4.0/SWATdata/extdata/plus_shapes/sub.shp"
```
These paths you can use to load the spatial data with your preferred method. The 
package sf (Pebesma, 2018; Pebesma et al., 2018) provides a very elegant way to 
work with spatial vector objects in R that have a direct integration for 
plotting in ggplot2 (Wickham, 2016).

```{r plot basin, echo=FALSE}
# Loading the sf and ggplot2 packages for the visualization
library(sf)
#> Linking to GEOS 3.6.2, GDAL 2.2.3, PROJ 4.9.3
library(ggplot2)
#> Registered S3 methods overwritten by 'ggplot2':
#>   method         from 
#>   [.quosures     rlang
#>   c.quosures     rlang
#>   print.quosures rlang

# Loading the spatial data as simple features
sub <- read_sf(sub_path)
riv <- read_sf(riv_path)

ggplot() +
  geom_sf(data = sub) +
  geom_sf(data = riv, col = "royalblue", lwd = 0.75) +
  geom_sf_label(data = riv, aes(label = Channel)) +
  theme_bw()

```

## First SWAT model runs

SWAT projects (e.g. one of the demo projects) can be executed from R with the 
functions run_swat2012() to run a SWAT2012 project or run_swatplus() to run a 
SWAT+ project. The minimum requirements to execute a SWAT project are to provide 
the project_path to the project folder on the hard drive and to define the 
simulation outputs that the simulation should return to R. The definition of the 
simulation output always follows a simple pattern. An output variable that 
should be returned must be defined using the function define_output(). Below is 
a simple example to run a SWAT+ project and return the variable = 'flo_out' that 
is written into the file = 'channel' file for the spatial unit = 1. Caution in 
the demo SWAT+ model setup the main channel is channel number 1.

The file structures for SWAT2012 and SWAT+ are different. The discharge in 
SWAT2012 is written in to the file “output.rch”. The discharge leaving a 
subbasin is written in the column “FLOW_OUT”.

```{r run SWAT2012}
q_sim_2012 <- run_swat2012(project_path = path_2012,
                           output = define_output(file = "rch",
                                                  variable = "FLOW_OUT",
                                                  unit = 1:3))                                  
```

```{r run SWAT+}
q_sim_plus <- run_swatplus(project_path = path_plus,
                           output = define_output(file = "channel",
                                                  variable = "flo_out",
                                                  unit = 1),
                           revision = 57)

#> Building 1 thread in 'Define:/your/path/swatplus_demo/.model_run':
#>  Thread 1 of 1   Time elapsed: 0S   Time remaining: 0S    
#>  Completed 1 thread in 0S
#> Performing 1 simulation on 1 core:
#>  Simulation 1 of 1   Time elapsed: 3S   Time remaining: 0S
#>  Completed 1 simulation in 3S
```

Running SWAT also provides information on the progress. This is very convenient 
for long simulation runs with many model evaluations. If you implement the SWAT 
function into other function calls the writing to the console can be an 
undesired behavior. You can quiet the function call with the argument 
quiet = TRUE.

## Output definition

The previous examples illustrated how to extract one variable (but maybe for 
several units) from the simulation outputs. To extract several variables you can 
simply combine them in a list(). In the example below we extract several water 
balance components of the SWAT+ demo simulations. Aggregated variables for a 
SWAT+ simulation are written to the ‘basin_wb’ file and there is only one 
spatial unit (the entire basin) written to that file. Therefore, the unit must 
be 1. Values different from 1 will not return any simulation results:

```{r ex SWAT+ sim}
wb_sim <- run_swatplus(project_path = path_plus,
           output = list(precip = define_output(file = "basin_wb",
                                                variable = "precip",
                                                unit = 1),
                         q_sur  = define_output(file = "basin_wb",
                                                variable = "surq_gen",
                                                unit = 1),
                         q_lat  = define_output(file = "basin_wb",
                                                variable = "latq",
                                                unit = 1),
                         eta    = define_output(file = "basin_wb",
                                                variable = "et",
                                                unit = 1)),
           revision = 57)

#> Building 1 thread in 'Define:/your/path/swatplus_demo/.model_run':
#>  Thread 1 of 1   Time elapsed: 1S   Time remaining: 0S    
#>  Completed 1 thread in 1S
#> Performing 1 simulation on 1 core:
#>  Simulation 1 of 1   Time elapsed: 2S   Time remaining: 0S
#>  Completed 1 simulation in 2S

```

## Exploring a simulation

The goal of SWATplusR is to return the SWAT simulations in a tidy format. The 
simulation results look slightly different if only one simulation was performed, 
parameters were modified in a run, or multiple parameter sets were used to 
perform a large number of model evaluations. In general the outputs follow a 
general structure.

If only one simulation was performed, where no model parameters were modified, 
run_swat*() returns a tibble (Müller and Wickham, 2019) with a date column and 
the columns with the simulated values of the extracted variables. The output 
table of the simulated water balance components has a date column and columns 
with the four extracted water balance components. The variable names are the 
individual names that we defined in the output definition:

In case no variable name is assigned in the output definition (as in the 
simulations of the discharge below), the variable name of the extracted variable 
is used as the variable name. When several units are defined in define_output() 
(e.g. in the example above unit = 1:3) a column in the output table is created 
for each unit and the unit number is added to the variable name. This is the 
case for the the simulation of ‘FLOW_OUT’ in the example above:

```{r explore sim 2012}
q_sim_2012
```


```{r explore sim+}
wb_sim
```

The tidy writing of the simulation results allows an easy processing of the data 
in the analysis workflow in R. Here is a quick example how to ‘wrangle’ the 
simulated discharge data with dplyr (Wickham et al., 2019) and tidyr (Wickham 
and Henry, 2018) and plotting it with ggplot2:

```{r tidy plot, echo = FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)

q_plot <- q_sim_2012 %>%
  select(date, FLOW_OUT_3) %>%
  rename(q_sim = FLOW_OUT_3) %>%
  left_join(., q_obs, by = "date") %>%
  rename(q_obs = discharge) %>%
  gather(., key = "variable", value = "discharge", -date)

ggplot(data = q_plot) +
  geom_line(aes(x = date, y = discharge, col = variable, lty = variable)) +
  scale_color_manual(values = c("black", "tomato3")) +
  theme_bw()
```

## Model parameter alteration

The capability to alter model parameters in a SWAT simulation is an essential 
option in SWATplusR. Model parameters can be modified in a SWAT simulation by 
passing the respective parameters with the argument parameter in run_swat*(). A 
single parameter set can be provided as a vector. Many parameter sets that 
should be used in simulations have to be provided in a tibble.

To modify a model parameter information on the type of change (relative, 
absolute change etc.), the extent of the change (e.g. all or selected subbasins/
land uses/ soil types etc.) and the magnitude of change are required. In 
SWATplusR most of that information is passed with the variable name. Thus, the 
variable names use a very specific syntax. A parameter name can consist of 
several parts, where some are required and others are optional. The minimum 
requirements for the elements of a parameter name are the actual name of the 
parameter in the model an the type of change. Options for the type of change are 
‘relchg’ and ‘pctchg’ to alter a parameter by a fraction or a percentage of the 
initial parameter value, “abschg” to add an absolute value to the initial 
parameter value, or “absval” to replace a parameter by an absolute value. In the 
example below the parameter curve number (‘cn2’) that is defined on HRU level 
will be increased by the absolute value 5 in all HRUs.

```{r change cn2}
par_mod <- c("cn2.hru|change = abschg" = 5)
```

Many options to constrain a parameter change are available in SWATplusR (in the 
current version of SWATplusR only for SWAT2012). Constraints for the alteration 
of a parameter is added with a ‘|’ and a logical expression. Below you find an 
example for the change of the curve number ‘CN2’ in a SWAT2012 setup, that is 
only effective for the subbasins 1 to 5 and only for the land uses “WWHT” and 
“CORN”. Also take care that ‘CN2’ is a parameter of the ‘.mgt’ input file:

```{r constrain params}
par_mod <- c("CN2.mgt|change = abschg | sub == 1:5 | luse %in% c('WWHT', 'CORN') " = 5)
```

Individual parameter names can be assigned with the :: operator. At the 
beginning of a parameter definition the individual parameter name is provided 
followed by ‘::’ and the actual parameter definition:

```{r change cn2}
par_mod <- c("my_name::cn2.hru|change = abschg" = 5)
```

## Simulation with a single parameter set

To define single parameter set a named vector is generated. The names define the 
parameters the type of change. Below a single parameter set is generated where 
all ‘cn2’ values in a SWAT+ model setup are reduced by 5 percent and the 
‘alpha.gw’ value was set to 0.5 globally.

```{r sim param set}
par_single <- c("cn2.hru|change = pctchg" = - 5,
                "alpha.gw|change = absval" = 0.5)
```

To include the parameter changes in the SWAT+ model execution the defined vector 
is simply passed with the argument parameter in the run_swatplus() function call:

```{r single param change}
q_sim <- run_swatplus(project_path = path_plus,
                      output = define_output(file = "channel",
                                             variable = "flo_out",
                                             unit = 1),
                      parameter = par_single,
                      revision = 57)

#> Building 1 thread in 'Define:/your/path/swatplus_demo/.model_run':
#>  Thread 1 of 1   Time elapsed: 1S   Time remaining: 0S    
#>  Completed 1 thread in 1S
#> Performing 1 simulation on 1 core:
#>  Simulation 1 of 1   Time elapsed: 2S   Time remaining: 0S
#>  Completed 1 simulation in 2S
```

## Simulations with many parameter sets

The same workflow works with several parameter sets when they are defined in a 
tibble. The example below uses the same parameters as above, but in this case a 
tibble was defined with 8 random realizations for each parameter within the 
defined boundaries.

```{r sims with multiple param sets}
par_set <- tibble("cn2.hru|change = abschg" = runif(8,-15,10),
                  "alpha.gw|change = absval" = runif(8, 0, 1))
par_set
```

The implementation in the run_swat*() function works the same way. To execute 
multiple SWAT simulations with many parameter sets the SWATplusR package 
provides an option parallel computing. To perform a parallel model execution the 
number of parallel threads is defined with the input argument n_thread.

```{r basin wb}
q_sim <- run_swatplus(project_path = path_plus,
                      output = list(q_sur  = define_output(file = "basin_wb",
                                                           variable = "surq_gen",
                                                           unit = 1),
                                    q_lat  = define_output(file = "basin_wb",
                                                           variable = "latq",
                                                           unit = 1)),
                      parameter = par_set,
                      n_thread = 4,
                      revision = 57)

#> Building 4 threads in 'Define:/your/path/swatplus_demo/.model_run':
#>  Thread 2 of 4   Time elapsed: 1S   Time remaining: 1S    
#>  Completed 4 thread in 2S
#> Performing 8 simulations on 4 cores:
#>  Simulation 4 of 8   Time elapsed: 4S   Time remaining: 4S
#>  Completed 8 simulations in 11S
```

## Exploring simulations with parameter alterations

When parameters were altered in a simulation run, the performed parameter 
changes are saved in the simulation outputs by default. As a consequence, the 
entire simulation output cannot be provided in a single tibble anymore and will 
be returned as a list with two main list entries. output$parameter provides all 
information on the parameter changes that were performed in the simulation runs. output$simulation provides the simulation results. The outputs of the previous 
model execution of the SWAT+ demo project has the following structure

The list entry $parameter holds two tibbles. The first tibble shows the values 
of the parameter changes of each model evaluation that was performed with 
run_swat*(). Each row of the tibble represents a parameter set that was used in 
a model run. The second tibble provides information on the parameter names, 
their actual names in the model, the type of change and the constraints on the 
parameter:

```{r explore more sims}
q_sim$parameter
#> $values
#> # A tibble: 8 x 2
#>        cn2 alpha
#>      <dbl> <dbl>
#> 1   0.360  0.401
#> 2 -11.3    0.151
#> 3 -11.1    0.752
#> 4  -5.67   0.608
#> 5  -9.94   0.590
#> 6   8.37   0.571
#> 7   4.52   0.408
#> 8   0.0230 0.115
#> 
#> $definition
#> # A tibble: 2 x 4
#>   par_name parameter file_name change
#>   <chr>    <chr>     <chr>     <chr> 
#> 1 cn2      cn2       hru       abschg
#> 2 alpha    alpha     gw        absval
```

The list entry $simulation provides the simulation runs that corresponds to each 
parameter set. Different to the single simulation without parameter changes is 
that now the simulated variables are not saved in a single tibble, but as a list 
of tibbles. In this case each simulated variable is saved in one tibble, where 
the first column of each tibble is the ‘date’ and the other columns are the 
individual simulation runs. Now the tibbles are named by the variables and each 
individual run saved in the tibbles is named ‘run_1’ to run ‘run_n’. In the 
example two variables were simulated, q_out$simulation$q_sur and 
q_out$simulation$q_lat. Each tibble holds the date and the runs 1 to 8:

```{r multiple sims}
q_sim$simulation
#> $q_sur
#> # A tibble: 3,653 x 9
#>    date       run_1 run_2 run_3 run_4 run_5 run_6 run_7 run_8
#>    <date>     <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
#>  1 2003-01-01 0.149 0.023 0.024 0.059 0.029 0.335 0.221 0.144
#>  2 2003-01-02 0.022 0.003 0.003 0.009 0.004 0.053 0.035 0.021
#>  3 2003-01-03 0.005 0.001 0.001 0.002 0.001 0.013 0.008 0.005
#>  4 2003-01-04 0.001 0     0     0.001 0     0.003 0.002 0.001
#>  5 2003-01-05 0     0     0     0     0     0.001 0.001 0    
#>  6 2003-01-06 0     0     0     0     0     0     0     0    
#>  7 2003-01-07 0     0     0     0     0     0     0     0    
#>  8 2003-01-08 0     0     0     0     0     0     0     0    
#>  9 2003-01-09 0     0     0     0     0     0     0     0    
#> 10 2003-01-10 0     0     0     0     0     0     0     0    
#> # … with 3,643 more rows
#> 
#> $q_lat
#> # A tibble: 3,653 x 9
#>    date       run_1 run_2 run_3 run_4 run_5 run_6 run_7 run_8
#>    <date>     <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
#>  1 2003-01-01 2.27  2.52  2.52  2.44  2.51  1.81  2.06  2.28 
#>  2 2003-01-02 1.67  1.9   1.90  1.82  1.89  1.32  1.51  1.68 
#>  3 2003-01-03 1.34  1.54  1.54  1.47  1.53  1.05  1.21  1.34 
#>  4 2003-01-04 1.14  1.32  1.32  1.26  1.31  0.894 1.03  1.15 
#>  5 2003-01-05 1.03  1.19  1.19  1.13  1.18  0.802 0.931 1.03 
#>  6 2003-01-06 0.956 1.10  1.10  1.05  1.10  0.744 0.866 0.961
#>  7 2003-01-07 0.906 1.04  1.04  0.994 1.03  0.704 0.822 0.911
#>  8 2003-01-08 0.869 0.995 0.994 0.952 0.988 0.675 0.788 0.874
#>  9 2003-01-09 0.838 0.957 0.956 0.917 0.95  0.651 0.761 0.843
#> 10 2003-01-10 0.812 0.925 0.925 0.888 0.919 0.63  0.737 0.817
#> # … with 3,643 more rows
```

## Saving and loading Simulations

In many cases it is advantageous to save the simulations incrementally after 
each model evaluation, for instance when long simulation runs would be lost if 
the computer or the R instance crashes, or when the simulation results cannot be 
stored in the RAM storage at all.

Simulations can be saved incrementally in an SQLite data base on the local hard 
drive, by passing a name for the save folder with the argument save_file in 
run_swat*(). By default the data base with the name save_file is saved in the 
project folder. An individual path for the save_file can be defined with the 
argument save_path. In case the simulations should be only saved in the data 
base and not returned back to R, the argument return_output = FALSE has to be 
set.

These arguments are simply added to the run_swatplus() function call to save the 
simulations in a data base instead of returning them back to R. Below the 
arguments are added to the previous simulation:

```{r save sims 1by1}
run_swatplus(project_path = path_plus,
            output = list(q_sur  = define_output(file = "basin_wb",
                                                 variable = "surq_gen",
                                                 unit = 1),
            q_lat  = define_output(file = "basin_wb",
                                   variable = "latq",
                                   unit = 1)),
            parameter = par_set,
            n_thread = 4,
            save_file = "q_sur_lat",
            return_output = FALSE,
            revision = 57)

#> Building 4 threads in 'Define:/your/path/swatplus_demo/.model_run':
#>  Thread 2 of 4   Time elapsed: 1S   Time remaining: 1S    
#>  Completed 4 thread in 2S
#> Performing 8 simulations on 4 cores:
#>  Simulation 4 of 8   Time elapsed: 3S   Time remaining: 3S
#>  Completed 8 simulations in 8S
```

You can see that the function did not return any results in R but generated a 
folder that we called “q_sur_lat” passed by the argument save_file in the 
project folder. That folder now contains several data bases. The first data 
base is called “par_dat.sqlite” and stores the parameter information and the 
date vector. Further, each of the parallel threads of the simulation generated 
a data base where each thread wrote the simulation results (This is necessary to 
avoid access conflicts when writing to data bases in parallel). The simulation 
data bases are always called “sim_thread_i.sqlite”, where “i” are the indexes of 
the individual threads (1 to 4 in this case).

SWATplusR provides two options to access the saved simulations. Loading large 
data bases into the R environment can take a long time. If you only want to know 
the the content of saved simulations (e.g. to check the successful simulation 
runs after a crash) you can scan the saved runs with the function 
scan_swat_run(). The function writes meta information on the saved SWAT 
simulations into the console. The function returns the following for our saved 
runs:

```{r access saved sims}
# Set the path to the folder where the date bases were saved
path_saved_q <- "Path:/to/the/folder/q_sur_lat"

scan_swat_run(save_dir = path_saved_q)
#> Simulation period:
#>  2003-01-01 to 2012-12-31
#>
#> Simulated variables:
#> q_lat: runs 1 to 8
#> q_sur: runs 1 to 8
#>
#> Parameter set:
#> # A tibble: 8 x 2
#>        cn2 alpha
#>      <dbl> <dbl>
#> 1   0.360  0.401
#> 2 -11.3    0.151
#> 3 -11.1    0.752
#> 4  -5.67   0.608
#> 5  -9.94   0.590
#> 6   8.37   0.571
#> 7   4.52   0.408
#> 8   0.0230 0.115
#> # A tibble: 2 x 4
#>   par_name parameter file_name change
#>   <chr>    <chr>     <chr>     <chr>
#> 1 cn2      cn2       hru       abschg
#> 2 alpha    alpha     gw        absval
```

scan_swat_run() returns information on the simulation period, which output variables were saved and which simulation runs were executed for these variables. Further, information on the used model parameter sets is written.

With load_swat_run() you can then load the simulations saved in the data bases. In case the entire set of simulations cannot be loaded in to the RAM storage the function provides the option to only load parts of the saved simulations. In the example below for example we load only the variable “q_lat” and only the first three runs into R:

```{r load sims}
q_subset <- load_swat_run(save_dir = path_saved_q,
                          variable = "q_lat",
                          run = 1:3)
```

```{r load sims2}
q_subset
#> $parameter
#> $parameter$values
#> # A tibble: 8 x 2
#>        cn2 alpha
#>      <dbl> <dbl>
#> 1   0.360  0.401
#> 2 -11.3    0.151
#> 3 -11.1    0.752
#> 4  -5.67   0.608
#> 5  -9.94   0.590
#> 6   8.37   0.571
#> 7   4.52   0.408
#> 8   0.0230 0.115
#> 
#> $parameter$definition
#> # A tibble: 2 x 4
#>   par_name parameter file_name change
#>   <chr>    <chr>     <chr>     <chr> 
#> 1 cn2      cn2       hru       abschg
#> 2 alpha    alpha     gw        absval
#> 
#> 
#> $simulation
#> $simulation$q_lat
#> # A tibble: 3,653 x 4
#>    date                run_1 run_2 run_3
#>    <dttm>              <dbl> <dbl> <dbl>
#>  1 2003-01-01 00:00:00 2.27  2.52  2.52 
#>  2 2003-01-02 00:00:00 1.67  1.9   1.90 
#>  3 2003-01-03 00:00:00 1.34  1.54  1.54 
#>  4 2003-01-04 00:00:00 1.14  1.32  1.32 
#>  5 2003-01-05 00:00:00 1.03  1.19  1.19 
#>  6 2003-01-06 00:00:00 0.956 1.10  1.10 
#>  7 2003-01-07 00:00:00 0.906 1.04  1.04 
#>  8 2003-01-08 00:00:00 0.869 0.995 0.994
#>  9 2003-01-09 00:00:00 0.838 0.957 0.956
#> 10 2003-01-10 00:00:00 0.812 0.925 0.925
#> # … with 3,643 more rows
```

## Further SWATplusR input arguments
The functions run_swat2012() and run_swatplus() have many further input 
arguments that allow to control important settings of a SWAT simulation. Here I 
will just mention a few of the essential ones. A detailed overview of the input 
arguments is provided in the respective function documentations. The simulation 
period can be controlled with the arguments start_date and end_date. The format 
writing the outputs of the simulations in the SWAT project folder is controlled 
by the arguments output_interval to define whether the simulations are written 
with daily, monthly or yearly time steps, years_skip to define the simulated 
years that should be skipped for writing the simulation outputs. With the 
argument run_index a subset of the provided parameter combinations can be 
simulated. This is useful, when for instance distributing the simulations on 
multiple computers, or when continuing the simulations after a crash.


