---
title: "03_param_opt"
author: "Danica Schaffer-Smith"
date: "5/14/2021"
output: html_document
---

## 03 - Parameter optimization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## R packages

There are many optimization routines available in R. The base R package stats (R Core Team, 2019) provides the function optim() that offers several standard methods for model parameter optimization. The hydromad (Andrews and Guillaume, 2018) package provides a complete environment for hydrological modeling in R. It also offers several optimization algorithms that are standard methods in hydrological modeling, such as the Shuffled Complex Evolution algorithm (SCE; Duan et al., 1993), or the Dynamically Dimensioned Search algorithm (DDS; Tolson and Shoemaker, 2007). A flexible integration of the DDS algorithm for R can also be found on the following github repository (Bass, 2014). Multiple goodness-of-fit functions are available from literature to evaluate simulated time series with observed time series of that variable. The hydroGOF package (Mauricio Zambrano-Bigiarini, 2017) summarizes frequently used functions for the evaluation of time series of hydrological variables.

# Package installation

If you do not have installed any of the required R package, follow the instructions for the respective R package.

The hydromad package can be installed following the instructions on the <hydromad>{http://hydromad.catchment.org/#installation} website.

The other R packages are available from CRAN and can be installed with the following commands:

```{r install pkg}
#install.packages("hydroGOF")
#install.packages("dplyr")
#install.packages("lubridate")
#install.packages("tidyr")
#install.packages("ggplot2")

```

# Loading R packages

```{r load pkg}
library(SWATplusR)
library(hydromad)
library(hydroGOF)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)

```

# Loading demo data

The optimization example uses the SWAT+ demo project available from SWATplusR. 
The demo project is a very simple model setups of a head watershed of the Little 
River Experimental Watershed (LREW; Bosch et al., 2007). You can load the to 
your hard drive as follows:

```{r load demo}
# The path where the SWAT demo project will be written
demo_path <- 'data_input/SWAT_demo'

# Loading the SWAT+ demo project on your hard drive
path_plus <- load_demo(dataset = "project",
                       version = "plus",
                       path = demo_path, 
                       revision = 57)
```

SWATplusR also provides observation data of daily discharge records at the main 
outlet of the demo for the time period 1968-01-01 until 2012-12-31. We will use 
the observation data to evaluate the model in each optimization step. The model 
will be evaluated for the time period 2003-01-01 to 2012-12-31. Therefore, we 
load the demo data and limit it to this time period:

```{r obs q}
q_obs <- load_demo(dataset = "observation")

q_obs <- filter(q_obs, date >= ymd("2003-01-01"),
                       date <= ymd("2012-12-31"))
```

## Definition of the function to optimize

Most optimizers require a function as an input argument that uses a parameter 
set as input and returns a scalar value as a result. To use the run_swatplus() 
function in the optimization we wrap a function around the SWAT model execution 
that will be implemented in the optimizer to search the optimum parameter set 
that minimizes the return value (minimization is usually the default option in 
optimizers). 

An important side note for running SWAT+, the discharge is written 
in ha⋅m⋅day−1 (in our case it is day!). A conversion to m3⋅s−1 requires to divide
by 8.64. Below you see how you can implement the model execution into a function 
that can be optimized:

```{r opt func}
swat_model <- function(par) {
  run_swatplus(project_path = path_plus,
               output = list(q_sim = define_output(file = "channel",
                                     variable = "flo_out",
                                     unit = 1)),
               parameter = par,
               start_date = "2000-01-01",
               end_date = "2012-12-31",
               years_skip = 3,
               quiet = TRUE,
               revision = 57)
}

swat_optim <- function(par, obs) {
  q_sim <- swat_model(par)
  nse_q <- - NSE(q_sim$simulation$q_sim/8.64, obs)
  return(nse_q)
  }
```

We defined a function that has one input argument par that is the named 
parameter vector that will be passed to run_swatplus(). The simulation will 
return the discharge at the main outlet (unit = 1) for the time period 
2003-01-01 to 2012-12-31 (considering the years_skip = 3 as warm up period). In 
this case we set the function to be quiet = TRUE. 

In a next step the function uses the observed discharge values from 
q_obs$discharge for the same period and evaluate the simulated discharges using 
the Nash Sutcliffe Efficiency criterion (Nash and Sutcliffe, 1970) that is 
available with the function NSE() from the package hydroGOF that is then 
returned as a single value. You can see that we implemented the negative NSE as 
we will minimize with optim()

## Parameter optimization with optim

The optim() function provides several optimization routines (see the ‘Details’ 
section in the R help file). Most of the offered algorithms, however, do not 
provide the option to define parameter boundaries. Though, constraining SWAT 
model parameters is essential as these are bound in the model. The 
method = 'L-BFGS-B' implements a ‘quasi-Newton’ method according to Byrd et al. 
(1995) and allows us to define parameter boundaries.

In the optimization example we will use 7 parameters that are frequently used 
for model calibration with respect to simulated discharge. The optim function 
requires starting values (par_init) and in the case of method = 'L-BFGS-B' we 
can also define upper (par_upr) and lower (par_lwr) boundaries for the 
parameters to optimize. We name the parameter sets using the syntax for SWAT 
model parameters that is required for the run_swat*() functions (see the Get 
started section on ‘Model parameter alteration’ to learn more on parameter 
names):

```{r set par} 
par_names <- c("cn2.hru | change = abschg",
               "lat_ttime.hru | change = absval",
               "lat_len.hru | change = absval",
               "k.sol | change = pctchg",
               "z.sol | change = pctchg",
               "epco.hru | change = absval",
               "esco.hru | change = absval")

par_init <- c(0, 3, 50, 0 , 0, 0.5, 0.5)
par_lwr  <- c(-15, 0.5,  10, -50, -50, 0, 0)
par_upr  <- c( 10,  50, 100,  50,  50, 1, 1)

names(par_init) <- par_names
names(par_lwr) <- par_names
names(par_upr) <- par_names

```

The defined optimization function swat_optim() can now be implemented in the 
algorithm as follows:

**THIS TAKES A WHILE TO RUN!**

```{r optim}
opt_bfgs <- optim(par = par_init, fn = swat_optim, method = "L-BFGS-B",
                  lower = par_lwr, upper = par_upr, obs = q_obs$discharge,
                  control = list(maxit = 100))
```

We omitted any feedback of the optimization in that case. If you want to get 
some feedback of the optimization routine you can set the control parameter 
trace = 6 in the list of control parameters above. The result of the 
optimization is saved in a list object. 

Relevant information is given by the list entries opt_bfgs$par that provides the 
final optimum parameter set and opt_bfgs$value that shows the optimized 
objective criterion, in our case, NSE:

```{r final par}
opt_bfgs$par
#      cn2.hru | change = abschg lat_ttime.hru | change = absval 
#                     0.69630817                      1.33555063 
#  lat_len.hru | change = absval         k.sol | change = pctchg 
#                    50.12257351                      0.70424355 
#        z.sol | change = pctchg      epco.hru | change = absval 
#                    -0.01189533                      0.99995859 
#     esco.hru | change = absval 
#                     0.27219482 

opt_bfgs$value
#> [1] -0.7130826
```

To visualize the discharge time series with the found parameter set we have to 
rerun the swat_model() we defined above with the optimum parameter set:

```{r optim swat run}
q_bfgs <- swat_model(opt_bfgs$par)
```

We will visualize the simulation with the optimized parameter set together with 
the observed discharge. We prepare our data with dplyr (Wickham et al., 2019) 
and tidyr (Wickham and Henry, 2018) and plot it with ggplot2:

```{r optim plot}
q_plot <- q_bfgs$simulation %>%
  mutate(q_sim = q_sim/8.64) %>%
  left_join(., q_obs, by = "date") %>%
  rename(q_obs = discharge) %>%
  gather(., key = "variable", value = "discharge", -date)

ggplot(data = q_plot) +
  geom_line(aes(x = date, y = discharge, col = variable, lty = variable)) +
  scale_color_manual(values = c("black", "tomato3")) +
  theme_bw()
```


## Parameter optimization with SCEoptim()

The implementation of the SCE algorithm provided by hydromad with the function 
SCEoptim() follows an almost identical syntax as the optim() function. 
Therefore, we can perform the optimization in almost the same manner as above. 
In SCEoptim() we will add however a few control arguments to define some 
settings in the optimization, such as the relative tolerance reltol = 0.001 that 
is used to stop the optimization when the improvements in NSE are below the 
threshold, tolsteps = 3 to stop after the third time the improvement in NSE was 
below the threshold, and trace = 1 to get feedback from each optimization cycle. 
The optimization run looks as follows:

```{r sceoptim}
opt_sce  <- SCEoptim(swat_optim, par_init, lower = par_lwr, upper = par_upr,
                     obs = q_obs$discharge,
                     control = list(reltol = 10^(-3), tolsteps = 3, trace = 1))
# Nr Iter  Nr Fun Eval    Current best function    Current worst function
#     1       408                 -0.73952                   1180.22
#     2       655                -0.751994                   1180.22
#     3       912                -0.753084                   1180.22
#     4      1135                 -0.75388                   1180.22
#     5      1336                -0.754416                   1180.22
#     6      1466                -0.755008                   1180.22
#Change in solution over [tolsteps] less than specified tolerance (reltol).
```

The optimization result with SCEoptim() provides more detailed information 
compared to optim(), such as the best parameter set of each iteration step. In 
our example you can see that all ten simulations actually resulted in almost the 
same model performance, a look at the respective parameter sets shows however, 
that some parameters strongly vary between these results. Therefore, it can be 
interesting to have a look on all ten simulation results. The best parameter set 
for each iteration step is saved in the variable opt_sce$BESTMEM.ALL. We will 
write them in a tibble (Müller and Wickham, 2019) to use them in our 
simulations. In a next step we can run the SWAT+ model with these ten parameter 
sets. We will use parallel computing here to have shorter computation times:

```{r SCEoptim par}
par_best <- as_tibble(opt_sce$BESTMEM.ALL)

q_sce <- run_swatplus(project_path = path_plus,
                      output = list(q_sim = define_output(file = "channel",
                                                          variable = "flo_out",
                                                          unit = 1)),
                      parameter = par_best,
                      start_date = "2000-01-01",
                      end_date = "2012-12-31",
                      years_skip = 3,
                      n_thread = 4,
                      revision = 57)

#Building 4 threads in 'data_input/SWAT_demo/swatplus_rev57_demo/.model_run': 
# Completed 4 threads in 1S                                                  
#Performing 6 simulations on 4 cores: 
# Completed 6 simulations in 8S 
```

To see the distribution of the parameters we first create dotty plots. To do 
that we have to add the optimized NSE values to the parameter table par_best. 
After that we again transform our data and plot with ggpot2:

```{r SCEoptim par plot}
par_plot <- q_sce$parameter$values %>%
  mutate(nse = apply(opt_sce$POP.FIT.ALL, 1, min),
         run = paste0("run_", sprintf("%02d", 1:6))) %>% # Vector length number sims
  gather(key = "par", value = "val", -nse, -run)


ggplot(par_plot) +
  geom_point(aes(x = val, y = abs(nse), col = run)) +
  facet_wrap(.~par, scales = "free") +
  theme_bw()

```

The time series of the simulated discharges are plotted similarly to the plot 
for the results of otim(). Though, some modifications were necessary in the data 
preparation and the plot command itself to plot all ten simulations:

```{r SCEoptim ts plot}
sce_plot <- q_sce$simulation$q_sim %>%
  mutate_if(is.numeric, list(~./8.64)) %>%
  gather(., key = "variable", value = "discharge", -date)

obs_plot <- rename(q_obs, q_obs = discharge)

ggplot() +
  geom_line(data = obs_plot, aes(x = date, y = q_obs), col = "black") +
  geom_line(data = sce_plot,
            aes(x = date, y = discharge, col = variable), alpha = 0.75) +
  scale_color_brewer(palette = "Paired") +
  ylab("discharge") +
  theme_bw()
```


## References

Andrews, F. and Guillaume, J.: Hydromad: Hydrological model assessment and development. [online] Available from: http://hydromad.catchment.org/ (Accessed 5 March 2019), 2018.

Bass, B.: Tolson & Shoemaker’s DDS Algorithm implementation in R. [online] Available from: https://github.com/bdb67/Dynamically-Dimensioned-Search (Accessed 19 March 2019), 2014.

Bosch, D. D., Sheridan, J. M., Lowrance, R. R., Hubbard, R. K., Strickland, T. C., Feyereisen, G. W. and Sullivan, D. G.: Little river experimental watershed database, Water Resources Research, 43(9), doi:10.1029/2006wr005844, 2007.

Byrd, R. H., Lu, P., Nocedal, J. and Zhu, C.: A limited memory algorithm for bound constrained optimization, SIAM Journal on Scientific Computing, 16(5), 1190–1208, doi:10.1137/0916069, 1995.

Duan, Q. Y., Gupta, V. K. and Sorooshian, S.: Shuffled complex evolution approach for effective and efficient global minimization, Journal of Optimization Theory and Applications, 76(3), 501–521, doi:10.1007/bf00939380, 1993.

Mauricio Zambrano-Bigiarini: HydroGOF: Goodness-of-fit functions for comparison of simulated and observed hydrological time series., 2017.

Müller, K. and Wickham, H.: Tibble: Simple data frames. [online] Available from: https://CRAN.R-project.org/package=tibble (Accessed 5 March 2019), 2019.

Nash, J. E. and Sutcliffe, J. V.: River flow forecasting through conceptual models part I - A discussion of principles, Journal of Hydrology, 10(3), 282–290, doi:10.1016/0022-1694(70)90255-6, 1970.

R Core Team: R: A language and environment for statistical computing., [online] Available from: https://www.r-project.org/ (Accessed 5 March 2019), 2019.

Tolson, B. A. and Shoemaker, C. A.: Dynamically dimensioned search algorithm for computationally efficient watershed model calibration, Water Resources Research, 43(1), doi:10.1029/2005wr004723, 2007.

Wickham, H. and Henry, L.: Tidyr: Easily tidy data with ’spread()’ and ’gather()’ functions. [online] Available from: https://CRAN.R-project.org/package=tidyr (Accessed 3 May 2019), 2018.

Wickham, H., François, R., Henry, L. and Müller, K.: Dplyr: A grammar of data manipulation. [online] Available from: https://CRAN.R-project.org/package=dplyr (Accessed 5 March 2019), 2019.